\blue{Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media - change the title to something close to the main topic of the manuscript ... something with viscous flow instability or viscous fingering)}

### 2017-02-02 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{section:Intro}
\medskip
Numerical investigation and modelling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Out of this mixed and complex formations, viscous flow instabilities can occur both within a single phase or when multiple phases flow through porous media. Forecasting the behaviour of multiphase flows or the future response of the system are major challenges that need to be taken into account. Injections of low-viscosity, high-density fluid $1$ (e.g. water) into a reservoir saturated with lighter but more viscous oil is an example where both viscous and gravitational instabilities may occur for two-phase immiscible and incompressible flow. These phenomena have great importance during the oil and gas recovery operations in both mature and new fields. Starting from oil and gas migration of contaminants through groundwater is the opposite problem governed by the same principles. The most complicated of processes include fingering that wil loccur in multiphase and compositional flow. Underground coal gasification is another potential application but more recently on CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture and storage (CCS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

@@

\subsection{Mathematical Formulation}\label{section:math-model}
\medskip
\KCnote{ - - to be putted in the mathematical formulation as intro - - 
Fingering will occur in miscible displacments when the viscocity ratio is greater than unity. As surface tension becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability by the basic mechanism associated with a less viscous fluid displacing a more viscous one. After a split, each of the new lobes of the finger is stable as a result of their being thinner than the finger from which they split. As a result of \textit{shielding}, one of these lobes will eventually outgrow the others and will then spread to occupy the appropriate width of the cell. In the process, it reaches a width that is again unstable to splitting, and the pattern repeats. Thus, surface tension plays an essential dual role; it must be weak enough for the tip front to be unstable, but it is also the physical force causing the \textit{spreading} and ensuing repeated branching. Important parameter is the mobility (or viscosity) ratio. This is given by Eqn.~\ref{eq:mobility-of-fluid} where $K$ is absolute permeability and $k_{r}$ is relative permeability :

@@

\noindent The two-phase flow through the porous media $\Omega$ may be described by the coupled extended Darcy and global mass conservation equations,
\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha},\;\;\; x_{i}\in\Omega, t>0 \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha},\;\;\; x_{i}\in\Omega, t>0\label{eqn:saturation_eqn}
\end{eqnarray}
where $S$, $\mu$, ${\bf K}$, $p$ and $\phi$ are the saturation, viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{k}$ is the saturation-weighted Darcy velocity of the $k$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ is saturation of the $\alpha$-phase with the constraint of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases.

\subsection{The Overlapping CVFEM formulation}
\medskip
Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equation (eq.~\ref{e:darcy_eq}). Hybrid (i.e., overlapping) basis functions are also developed to allow CV-based velocity to be extrapolated across the entire element. $\underline{\underline{\sigma}}_{k}=\underline{\underline{\sigma}}_{k}\left({\mathbf K}, \mathcal{K}_{rk}, S_{K}, \mu_{k}\right)$ is an absorption-like term that lies in both CV and FEM spaces. Though, whilst the saturation field is calculated in CV space, permeability $\left({\mathbf K}\right)$ is piecewise constant in FE space to allow efficient representation of surface parametrised geometries. The global saturation equation (Eqn.~\ref{eqn:saturation_eqn}) is discretised in space with CV basis function and with the $\theta$-method in time as described by \citet{gomes_book_2012}. Velocities across CV interfaces (within and between elements, see fig.\ref{fig:fem_cv} below) are calculated through a directional-weighted flux-limited scheme based on upwind value of $\sigma$ at individual CV as described by \citet{gomes_2013}~\citep[see also][]{jackson_2013}.

@@

where $\Omega_{CVi}$ and $\Gamma_{CVi}$ are the volume and boundary of CV i respectively, n is normal vector of the CV and $\theta$ varies smoothly between $0.5$ (corresponding to CN) and $1$ (corresponding to BE) to avoid the introduction of spurious oscillations for large grid Courant numbers. The Eq.\ref{global_mass_balance} can be expressed in amtrix form at time level $n+1$ as
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  
\section{Brief Overview of Viscous Instabilities}\label{section:ViscousInstabilities}
%\textcolor{red}{KC: could you add 1-2 short paragraphs on viscous instabilities ... focus on the phenomena rather than a literature review, that may take place in the Intro Section}
As in many areas of fluid mechanics, the behaviour of the flow for conditions that exceed critical limits, are those that occur under highly critical conditions. Under this assumption, viscous fingering is no exception. It is a viscous driven hydrodynamic instability that results the collapse of the uniform interface between the fluids of different viscosity. This type of instability can be found across many research areas, disciplines and scales. Starting from small scale flow instabilities in Hele-Shaw cells that can provide the basic mathematical model, to coupled chromatographic separation systems for chemical and medical applications and reaching up to large scale reservoir systems.

@@

First, how satisfactory are the effects of modelled-observed viscous fingering on displacement performance? Second, how is a viscous instability affected by parameters including mobility ratio and/or permeability variance. \textcolor{red}{ KC: something came to me right now, And third, can the viscous fingering be scaled?}

\subsection{Designing the Numerical Experiment}\label{section:NumericalExperiment}
\textcolor{red}{WR: could add here a paragraph describing the original geometry and simulation set up (initial and boundary conditions)and maybe a table with the differences between the diff numerical experiments.}

@@

\begin{equation}
M_{\sigma} {{\underline{u}}^{n+1}} = C {\underline{p}^{n+1}} + {\underline{s}_u ^{n+1}}
\label{mass balance eq}
\end{equation}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:Intro}
\medskip
Numerical investigation and modelling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Out of this mixed and complex formations, viscous flow instabilities can occur both within a single phase or when multiple phases flow through porous media. Forecasting the behaviour of multiphase flows or the future response of the system are major challenges that need to be taken into account. Injections of low-viscosity, high-density fluid $1$ (e.g. water) into a reservoir saturated with lighter but more viscous oil is an example where both viscous and gravitational instabilities may occur for two-phase immiscible and incompressible flow. These phenomena have great importance during the oil and gas recovery operations in both mature and new fields. Starting from oil and gas migration of contaminants through groundwater is the opposite problem governed by the same principles. The most complicated of processes include fingering that wil loccur in multiphase and compositional flow. Underground coal gasification is another potential application but more recently on CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture and storage (CCS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

@@

@@

\noindent The two-phase flow through the porous media $\Omega$ may be described by the coupled extended Darcy and global mass conservation equations,
\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha},\;\;\; x_{i}\in\Omega, t>0 \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha},\;\;\; x_{i}\in\Omega, t>0\label{eqn:saturation_eqn}
\end{eqnarray}
where $S$, $\mu$, ${\bf K}$, $p$ and $\phi$ are the saturation, viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{k}$ is the saturation-weighted Darcy velocity of the $k$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ is saturation of the $\alpha$-phase with the constraint of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases.
The main equations that will be used along with the following OCVFEM techique can be summarised below:

\begin{equation}
M_{\sigma} {{\underline{u}}^{n+1}} = C {\underline{p}^{n+1}} + {\underline{s}_u ^{n+1}}
\label{e:mas_bal}
\end{equation}

\begin{equation}
B^T {{\underline{u}}^{n+1}} = {\underline{s}_p ^{n+1}}
\label{e:force_bal}
\end{equation}

\begin{equation}
B^T M_{\sigma} ^{-1} C{{\underline{p}}^{n+1}} = {\underline{s}_p ^{n+1}} - B^T M_{\sigma} ^{-1} {{\underline{s}}^{n+1}} 
\label{e:pressure_eq}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{The Overlapping CVFEM formulation}
\medskip
Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equation (eq.~\ref{e:darcy_eq}). Hybrid (i.e., overlapping) basis functions are also developed to allow CV-based velocity to be extrapolated across the entire element. $\underline{\underline{\sigma}}_{k}=\underline{\underline{\sigma}}_{k}\left({\mathbf K}, \mathcal{K}_{rk}, S_{K}, \mu_{k}\right)$ is an absorption-like term that lies in both CV and FEM spaces. Though, whilst the saturation field is calculated in CV space, permeability $\left({\mathbf K}\right)$ is piecewise constant in FE space to allow efficient representation of surface parametrised geometries. The global saturation equation (Eqn.~\ref{eqn:saturation_eqn}) is discretised in space with CV basis function and with the $\theta$-method in time as described by \citet{gomes_book_2012}. Velocities across CV interfaces (within and between elements, see fig.\ref{fig:fem_cv} below) are calculated through a directional-weighted flux-limited scheme based on upwind value of $\sigma$ at individual CV as described by \citet{gomes_2013}~\citep[see also][]{jackson_2013}.

@@

where $\Omega_{CVi}$ and $\Gamma_{CVi}$ are the volume and boundary of CV i respectively, n is normal vector of the CV and $\theta$ varies smoothly between $0.5$ (corresponding to CN) and $1$ (corresponding to BE) to avoid the introduction of spurious oscillations for large grid Courant numbers. The Eq.\ref{global_mass_balance} can be expressed in amtrix form at time level $n+1$ as

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Overview of Viscous Instabilities}\label{section:ViscousInstabilities}

\KCnote{ - - to be putted in the mathematical formulation as intro - - 
Fingering will occur in miscible displacments when the viscocity ratio is greater than unity. As surface tension becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability by the basic mechanism associated with a less viscous fluid displacing a more viscous one. After a split, each of the new lobes of the finger is stable as a result of their being thinner than the finger from which they split. As a result of \textit{shielding}, one of these lobes will eventually outgrow the others and will then spread to occupy the appropriate width of the cell. In the process, it reaches a width that is again unstable to splitting, and the pattern repeats. Thus, surface tension plays an essential dual role; it must be weak enough for the tip front to be unstable, but it is also the physical force causing the \textit{spreading} and ensuing repeated branching. Important parameter is the mobility (or viscosity) ratio. This is given by Eqn.~\ref{eq:mobility-of-fluid} where $K$ is absolute permeability and $k_{r}$ is relative permeability :

As in many areas of fluid mechanics, the behaviour of the flow for conditions that exceed critical limits, are those that occur under highly critical conditions. Under this assumption, viscous fingering is no exception. It is a viscous driven hydrodynamic instability that results the collapse of the uniform interface between the fluids of different viscosity. This type of instability can be found across many research areas, disciplines and scales. Starting from small scale flow instabilities in Hele-Shaw cells that can provide the basic mathematical model, to coupled chromatographic separation systems for chemical and medical applications and reaching up to large scale reservoir systems.

@@

@@

\begin{equation}
M_{\sigma} {{\underline{u}}^{n+1}} = C {\underline{p}^{n+1}} + {\underline{s}_u ^{n+1}}
\label{mas_bal}
\end{equation}
